---
title: "make_per_sample_metadata"
output: html_document
---
```{r}
library(googledrive)
library(tidyverse)
```


# Get the transcribed sample data (documents from initial 4mm sieving of each sample, in the lab or field) from google drive 
```{r}

name_of_transcription_file_on_drive = "X_sample_data_transcribed" # copy the file's name exactly as it appears on google drive

dribble = drive_get(name_of_transcription_file_on_drive, shared_drive = 'Bogar Lab') # get necesary data for downloading

local_path <- paste0("Inputs/", 
                     name_of_transcription_file_on_drive, 
                     "_downloaded-", 
                     format(Sys.Date(), "%Y%m%d"), 
                     ".csv") #define what this local file name will be, with today's date

drive_download(dribble, type = "csv", path = local_path, overwrite = TRUE)

x_transcribed_data = read_csv(local_path, col_names = FALSE)

```

```{r}

name_of_transcription_file_on_drive = "YZ_sample_data_transcribed" # copy the file's name exactly as it appears on google drive

dribble = drive_get(name_of_transcription_file_on_drive, shared_drive = 'Bogar Lab') # get necesary data for downloading

local_path <- paste0("Inputs/", 
                     name_of_transcription_file_on_drive, 
                     "_downloaded-", 
                     format(Sys.Date(), "%Y%m%d"), 
                     ".csv") #define what this local file name will be, with today's date

drive_download(dribble, type = "csv", path = local_path, overwrite = TRUE)

yz_transcribed_data = read_csv(local_path, col_names = FALSE)

```

# Re-orient data tables

Rename the columns with their Sample_IDs:
```{r tidy}
# x samples
sampleIDrow = which(x_transcribed_data[[1]] == "Sample_ID") # find which row # contains all the Sample_IDs
colnames(x_transcribed_data) = as.character(x_transcribed_data[sampleIDrow,]) # make Sample_IDs the column names
x_transcribed_data = x_transcribed_data %>%
  rename(futureColumn = Sample_ID) %>% # rename the first column, which actually contains all the future column names (variable names? measurement names?)
  dplyr::slice(-sampleIDrow) # important to take the sample_ID row out so pivoting doesn't get confused later
head(x_transcribed_data)

# y and z samples
sampleIDrow = which(yz_transcribed_data[[1]] == "Sample_ID")
colnames(yz_transcribed_data) = as.character(yz_transcribed_data[sampleIDrow,])
yz_transcribed_data = yz_transcribed_data %>%
  rename(futureColumn = Sample_ID) %>%
  dplyr::slice(-sampleIDrow)

head(yz_transcribed_data)
```

Pivot to get sample_ID as its own column and all the measurements in rows
```{r}
x_transcribed_data_pivoted = x_transcribed_data %>%
  pivot_longer(cols = !futureColumn, 
               names_to = "Sample_ID", 
               values_to = "value"
               ) %>% # pivot longer first, making a column called Sample_ID
  pivot_wider(names_from = futureColumn, 
              values_from = value) # then wider, making each variable its own column

head(x_transcribed_data_pivoted)

yz_transcribed_data_pivoted = yz_transcribed_data %>%
  pivot_longer(cols = !futureColumn, 
               names_to = "Sample_ID", 
               values_to = "value"
               ) %>% # pivot longer first, making a column called Sample_ID
  pivot_wider(names_from = futureColumn, 
              values_from = value) # then wider, making each variable its own column

head(yz_transcribed_data_pivoted)
```
Check which column names are compatible
```{r}
# which column names in the x sample data sheet weren't in the yz sample data sheet?
colnames(x_transcribed_data_pivoted)[which(!(colnames(x_transcribed_data_pivoted) %in% colnames(yz_transcribed_data_pivoted)))]

print("-----------------------------------------------
      ------------------------------------------------")

# which column names in the yz sample data sheet weren't in the x sample data sheet?
colnames(yz_transcribed_data_pivoted)[which(!(colnames(yz_transcribed_data_pivoted) %in% colnames(x_transcribed_data_pivoted)))]

```

Resolve column name incompatibilities
```{r resolve xyz colnames}
#Sampling_date and sieve_date -- for x samples, sampling date is Sieve_date. For yz samples, Sampling_date is not recorded in this sheet
x_transcribed_data_colnamefix = x_transcribed_data_pivoted %>% 
  mutate(sieve_date = Sampling_date)

yz_transcribed_data_colnamefix = yz_transcribed_data_pivoted %>% 
  mutate(Sampling_date = "fill_in_from_core_data_later")

# Site is also not currently recorded in this sheet for yz samples
yz_transcribed_data_colnamefix = yz_transcribed_data_colnamefix %>% 
  mutate(Site = "fill_in_from_core_data_later")

#Equip_mgr_initials not relevant for yz samples
yz_transcribed_data_colnamefix = yz_transcribed_data_colnamefix %>% 
  mutate(Equip_mgr_initials = "NA_to_YZ_samples")

#Lroot and rock masses have _g suffix on yz only
x_transcribed_data_colnamefix = x_transcribed_data_colnamefix %>% 
  rename(Lroot_empty_mass_g = Lroot_empty_mass) %>%
  rename(rock_empty_mass_g = rock_empty_mass)

#Lroot and rock pictures by same person always for x samples, sometimes different person for yz samples
x_transcribed_data_colnamefix = x_transcribed_data_colnamefix %>% 
  mutate(Lroot_photo_by = Lroot_and_rock_photos_by) %>%
  mutate(rock_photo_by = Lroot_and_rock_photos_by) %>%
  select(!Lroot_and_rock_photos_by)

#Core_ID missing from yz samples; we need core data to add this 
yz_transcribed_data_colnamefix = yz_transcribed_data_colnamefix %>% 
  mutate(Core_ID = "fill_in_from_core_data_later")

# fraction column has longer name, in yz sample data, won't always be present
yz_transcribed_data_colnamefix = yz_transcribed_data_colnamefix %>% 
  rename(fraction = `fraction (if written on sample tube)`) %>%
  mutate(fraction = case_when(
    fraction != NA ~ fraction,
    fraction == NA ~ "fill_in_from_core_data_later"
    )
  )

#ROOTS_W_and_ H_freeze_time (x) vs all_ROOTS_freeze_time (yz); W and H were the only kinds of roots we froze for x samples, so these can be the same column
x_transcribed_data_colnamefix = x_transcribed_data_colnamefix %>% 
  rename(all_ROOTS_freeze_time = `ROOTS_W_and_ H_freeze_time`) 

# "RNA_A_freeze_time"                  "RNA_B_and_C_freeze_time"             "S_FM_mass_g"  ; all of these simply were recorded for X samples and not for YZ samples
yz_transcribed_data_colnamefix = yz_transcribed_data_colnamefix %>% 
  mutate(RNA_A_freeze_time = "NA_to_YZ_samples") %>% 
  mutate(RNA_B_and_C_freeze_time = "NA_to_YZ_samples") %>% 
  mutate(S_FM_mass_g = "NA_to_YZ_samples")

# notes columns are messy but I'm leaving them alone for now

# "sieving_start_time"                     "sieving_end_time" are in YZ but not X data; for X data, the recorded subsampling_start_time was effectively the sieving start time. We didn't record a sieving end time, but everything_cool_time is certianly after the sieving ended. I'm not going to change these, because while they approximate each other they're not exactly the same and I don't know that we need to know sieving start and end time exactly. 
x_transcribed_data_colnamefix = x_transcribed_data_colnamefix %>% 
  mutate(sieving_start_time = "after_subsampling_start") %>% 
  mutate(sieving_end_time = "before_everything_cool")

# mass of soil for pH was approximated by volume in the field, not measured
x_transcribed_data_colnamefix = x_transcribed_data_colnamefix %>% 
  mutate(S_pH_mass_g = "volume_approximated_in_field")

#"small_tin_mass_full_g"                  "large_tin_mass_full_g"                 "all_set_to_dry_time" Not recorded in the field on X data sheets; these samples were set out to air dry by Glade(?) after I left for SSC 205, so I need to track down these data elsewhere
x_transcribed_data_colnamefix = x_transcribed_data_colnamefix %>% 
  mutate(small_tin_mass_full_g = "recorded_elsewhere_June2026?") %>%
  mutate(small_tin_mass_empty_g = "recorded_elsewhere_June2026?") %>%
  mutate(large_tin_mass_full_g = "recorded_elsewhere_June2026?") %>%
  mutate(all_set_to_dry_time = "NA_to_X_samples")

# we weren't documenting animal burrows explicitly in the field
x_transcribed_data_colnamefix = x_transcribed_data_colnamefix %>% 
  mutate(`Animal_burrow_(Y/N)` = "not_recorded_for_X_samples")

# pre and post trim photos only taken systematically for yz samples; best approximate for x samples would be core section photos. + other tube removal types of data we didn't record for x samples
x_transcribed_data_colnamefix = x_transcribed_data_colnamefix %>% 
  mutate(pre_trim_photo_by = "see_core_section_photos_from_field") %>%
  mutate(post_trim_photo_by = "see_core_section_photos_from_field") %>%
  mutate(removed_soil_mass_g = "not_recorded_for_X_samples") %>%
  mutate(tube_length_cm = "not_recorded_for_X_samples") %>%
  mutate(`other_info_on_sample_tube (if present)` = "not_recorded_for_X_samples")
  

# ok now I'm fixing the notes

yz_transcribed_data_colnamefix = yz_transcribed_data_colnamefix %>% 
  rename(notes_during_collection_narritive = Notes_during_collection) %>% 
  rename(notes_during_transcription_EF = Notes_during_transcription) %>% 
  mutate(notes_during_transcription_AG = "NA_to_YZ_samples") %>%
  mutate(notes_during_collection_annotations = "included_in_narritive_for_YZ_samples")

x_transcribed_data_colnamefix = x_transcribed_data_colnamefix %>% 
  select(!page_on_pdf_scan)

```

```{r}
# which column names in the x sample data sheet weren't in the yz sample data sheet?
colnames(x_transcribed_data_colnamefix)[which(!(colnames(x_transcribed_data_colnamefix) %in% colnames(yz_transcribed_data_colnamefix)))]

print("-----------------------------------------------
      ------------------------------------------------")

# which column names in the yz sample data sheet weren't in the x sample data sheet?
colnames(yz_transcribed_data_colnamefix)[which(!(colnames(yz_transcribed_data_colnamefix) %in% colnames(x_transcribed_data_colnamefix)))]

#ok cool only the notes are different now


```

# Join x and yz sample data
```{r join x and yz sample data}

xyz_sample_data = rbind(x_transcribed_data_colnamefix, yz_transcribed_data_colnamefix)
xyz_sample_data
```

export data for all june samples from initial subsampling pushes!
```{r}

local_path_tidydata <- paste0("Results/", 
                     "june_subsampling_metadata_xyz-", 
                     format(Sys.Date(), "%Y%m%d"), 
                     ".csv") #define what this local file name will be, with today's date

write_csv(xyz_sample_data, local_path_tidydata)
```

wahooo!

# get Core_ID and fraction from core data
Do this better from complete transcribed core data later. As of 14 Aug 2026, I'm using this partial transcription I did in June:
```{r}

name_of_transcription_file_on_drive = "core_data_transcribed_20260620" # copy the file's name exactly as it appears on google drive

dribble = drive_get(name_of_transcription_file_on_drive, shared_drive = 'Bogar Lab') # get necesary data for downloading

local_path <- paste0("Inputs/", 
                     name_of_transcription_file_on_drive, 
                     "_downloaded-", 
                     format(Sys.Date(), "%Y%m%d"), 
                     ".csv") #define what this local file name will be, with today's date

drive_download(dribble, type = "csv", path = local_path, overwrite = TRUE)

partial_transcribed_june_core_data = read_csv(local_path)

```
remove rows not containing real samples we collected
```{r}
core_IDs_fractions = partial_transcribed_june_core_data %>%
  filter(startsWith(Sample_ID, "DSC"))
```

break 016 into 016A and 016B
```{r}

sixteenIndex = which(core_IDs_fractions$Sample_ID == "DSC_016")
sixteenRowA = core_IDs_fractions[sixteenIndex,] %>%
  mutate(Sample_ID = "DSC_016A")

sixteenRowB = core_IDs_fractions[sixteenIndex,] %>%
  mutate(Sample_ID = "DSC_016B")

core_IDs_fractions_16s_fixed = core_IDs_fractions %>%
  rbind(sixteenRowA) %>%
  rbind(sixteenRowB) %>%
  filter(Sample_ID != "DSC_016")
```

join to metadata
```{r}
metadata_add_coreinfo = xyz_sample_data %>%
  select(!c(Core_ID, fraction, Site)) %>%
  full_join(core_IDs_fractions_16s_fixed, by = "Sample_ID")
```

add a new variable called "microsite" to link tree cores to their non-tree cores
```{r}
metadata_add_coreinfo = metadata_add_coreinfo %>%
  mutate(microsite = case_when(
    (Core_ID == "B-E" | Core_ID == "B-R") ~ "S1",
    (Core_ID == "B-F" | Core_ID == "B-S") ~ "S2",
    (Core_ID == "B-H" | Core_ID == "B-U") ~ "S3",
    (Core_ID == "U-M" | Core_ID == "U-Z") ~ "N1",
    (Core_ID == "U-J" | Core_ID == "U-W") ~ "N2",
    (Core_ID == "U-01" | Core_ID == "U-02") ~ "N3",
    .default = NA
  ))
```

add a new variable called "tnt" for "tree/no tree"
```{r}
metadata_add_coreinfo = metadata_add_coreinfo %>%
  mutate(tnt = case_when(
    Core_ID == "B-E" | 
       Core_ID == "B-F" |
       Core_ID == "B-H" |
       Core_ID == "U-M" |
       Core_ID == "U-J" |
       Core_ID == "U-01" ~ "tree",
     Core_ID == "B-R" |
       Core_ID == "B-S" |
       Core_ID == "B-U" |
       Core_ID == "U-Z" | 
       Core_ID == "U-W" |
       Core_ID == "U-02" ~ "no_tree",
    .default = NA
  ))
```

parse fractions
```{r}
metadata_add_coreinfo <- metadata_add_coreinfo %>%
  mutate(
    sample_bottom_depth = case_when(
      fraction == "O" ~ 0,
      TRUE ~ as.numeric(str_extract(fraction, "(?<=-)\\d+"))
    ),
    sample_top_depth = case_when( # FIX THIS when we have good core data; ideally we'd use the o cm thickness to make top depth for Os negative
      fraction == "O" ~ 0,
      TRUE ~ as.numeric(str_extract(fraction, "^\\d+"))
    )
  )
```

# Add moisture data
```{r}
od_moisture_tidydata = read_csv("../../Moisture/Results/od_moisture_tidied-20260814.csv")

# also read in ad moisture data when it's tidied! not yet tidied as of 14 Aug 2026
```

```{r}
metadata_add_odm = metadata_add_coreinfo %>%
  left_join(od_moisture_tidydata, by = "Sample_ID")
```


# Add pH data
```{r}
pH_tidydata = read_csv("../../pH/Results/pH_tidied-20260814.csv")
```

```{r}

metadata_add_pH = metadata_add_odm %>%
  left_join(pH_tidydata, by = "Sample_ID")
```

# Add texture data

```{r}
kimber_texture = read_csv("../../../Data/Texture/Bogar_Texture.csv")

tex = kimber_texture %>%
  rename(Sample_ID = long_sample_id) %>%
  mutate(Sample_ID = stringr::str_remove_all(Sample_ID, "\\s"))

metadata_add_tex = metadata_add_pH %>%
  left_join(tex, by = "Sample_ID")
```


export most complete metadata csv
```{r}

local_path_tidydata <- paste0("Results/", 
                     "DSC_per_sample_metadata_compiled-", 
                     format(Sys.Date(), "%Y%m%d"), 
                     ".csv") #define what this local file name will be, with today's date

write_csv(metadata_add_tex, local_path_tidydata)
```


